---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
set.seed(1)
```

# subnetr

<!-- badges: start -->
[![R-CMD-check](https://github.com/xavienzo/subnetr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/xavienzo/subnetr/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`subnetr` extracts covariate-related subnetworks from whole-brain connectome
data and provides simulation-based power analysis for study design. It
implements the subnetwork detection framework of Chen et al. (2023), which
builds on the adaptive dense subgraph discovery model of Wu et al. (2022).

A covariate is typically related to a number of edges connecting multiple brain
areas in an organized structure, but neither the covariate-related edges nor
that structure is known in advance. Mass-univariate analysis applies a single
threshold to every edge and returns a set of unrelated significant edges,
recognizing no network topology; it also incurs a severe multiple-comparison
burden. Subnetwork detection instead identifies node sets whose mutual
connections are jointly associated with the covariate. Edge-wise test
statistics are screened into a sparse weighted graph, an adaptive density
objective is maximized by a greedy algorithm to extract candidate subnetworks,
and a permutation test assigns each of them a p-value controlling the
family-wise error rate.

The extraction and permutation machinery is implemented in C++, which makes it
practical to refit the procedure thousands of times and therefore to estimate
power by simulation.

## Installation

```r
# install.packages("remotes")
remotes::install_github("xavienzo/subnetr")
```

## Basic analysis

The input is subject-level connectivity: an `n_subject` by `n_edge` matrix of
vectorized connectomes, together with a predictor and any nuisance covariates.

```{r quickstart, message = FALSE}
library(subnetr)

sim <- simulate_fc(n = 120, n_nodes = 80, cluster_size = 20,
                   f2 = 0.06, seed = 42)

W   <- sim$W                       # -log10 p per edge, from edge_stats()
fit <- subnet(W, n_perm = 999, seed = 1)

fit
```

For observed data, the first two lines are replaced by

```r
W <- edge_stats(fc, x = predictor, covariates = cbind(age, sex, motion))
```

where `fc` is either an `n_subject` × `n_edge` matrix or an
`n_node` × `n_node` × `n_subject` array. Each edge is regressed on the
predictor with the covariates partialled out, and the returned matrix holds the
resulting edge-wise evidence.

The reported p-values are already adjusted for multiplicity across all
extracted subnetworks and should not be adjusted again.

### Visualizing the result

```{r heatmap, fig.height = 4.4, fig.width = 10, fig.alt = "Left: connectivity matrix in its original node order, showing no visible structure. Right: the same matrix reordered so the detected subnetwork forms an outlined block on the diagonal."}
plot(fit, what = "both")
```

The left panel shows the adjacency matrix in the original node order, in which
the subnetwork is distributed across arbitrary indices and not apparent. The
right panel shows the same matrix with nodes reordered so that each extracted
subnetwork occupies a contiguous diagonal block; significant blocks are
outlined. Both panels use a common colour scale, labelled with the statistic
displayed — here `-log10(p)`, since that is what `edge_stats()`
returned. Use `weight_label` to set the label for matrices constructed by other
means.

Both panels are worth inspecting. Reordering is a permutation of node labels,
and a permutation alone can impose apparent structure on noise; the comparison
between panels indicates whether the block is substantive. The permutation test
formalizes that comparison.

Node membership is available directly:

```{r members}
fit$subnetworks[[1]]        # node indices of the leading subnetwork
table(membership(fit))      # 0 = background
```

### Multiple subnetworks

The method does not assume a single subnetwork. In the following example two
subnetworks of different size and effect size are planted, and both are
recovered as separate blocks.

```{r multi, message = FALSE}
sim2 <- simulate_fc(n = 150, n_nodes = 90, cluster_size = c(20, 12),
                    f2 = c(0.08, 0.14), seed = 4)
fit2 <- subnet(sim2$W, n_perm = 999, seed = 2)

res <- as.data.frame(fit2)[, c("subnet", "size", "density", "p_value",
                               "significant")]
res$density <- round(res$density, 2)
res
```

```{r multi-heatmap, fig.height = 4.4, fig.width = 10, fig.alt = "Left: connectivity matrix with two planted subnetworks in the original node order, showing no visible structure. Right: the same matrix reordered, with the two detected subnetworks outlined as separate diagonal blocks of different size."}
plot(fit2, what = "both")
```

Extracted subnetworks are disjoint by construction: each peeling round passes
the nodes it removes to the next round, so every node belongs to exactly one
block. The p-values are corrected jointly across all blocks returned, whether
or not they reach significance, because each permutation repeats the entire
extraction and contributes only its most extreme block. Reporting two
significant subnetworks among seven extracted therefore incurs no additional
multiplicity penalty.

Two qualifications apply. The null edge probability is estimated over the whole
graph and so includes edges carrying signal, which renders the test
conservative when one subnetwork is considerably stronger than another. A
partition also cannot represent *overlapping* subnetworks: a node belonging to
two systems is assigned to whichever block claims it first. See
`vignette("multiple-subnetworks")`.

## Power analysis

`power_curve()` estimates power by simulating the complete procedure,
permutation test included, over a grid of sample sizes.

```{r power, message = FALSE, fig.height = 4.4, fig.width = 6.5, fig.alt = "Power curves against sample size, rising from near zero to one, for any detection and for correct recovery of the planted subnetwork"}
pw <- power_curve(n = c(30, 60, 100, 150, 220),
                  n_nodes = 80, cluster_size = 15, f2 = 0.025,
                  n_sim = 200, n_perm = 199, seed = 7, progress = FALSE)

pw
plot(pw, target = 0.8)
```

Two quantities are reported. **Power** is the probability of declaring any
subnetwork significant. **Recovery** is the probability of identifying one that
overlaps the true subnetwork with Sørensen–Dice coefficient at least
`dice_cut`. Recovery is the more stringent criterion, and the appropriate one
when the inferential claim concerns which nodes are involved.

```{r requiredn}
required_n(pw, target = 0.8)
```

Effect sizes are specified as Cohen's `f^2` per edge. The argument
`rho_in` gives the proportion of within-subnetwork edges that carry the effect;
the default of 0.9 reflects the expectation that a subnetwork is not uniformly
affected.

## Method

**Edge-wise inference.** Each edge of the connectome is fitted by a general
linear model of the connectivity measure on the covariate of interest,
adjusting for nuisance covariates. The inferential results are stored in a
matrix `W = {w_ij}` with `w_ij = -log10(p_ij)`, which forms the input to
subnetwork extraction. Any valid statistical model producing such a matrix may
be used.

**Screening.** `W` is thresholded at a value `r`, edges below it being set to
zero. What remains is a sparse weighted graph of suprathreshold edges.

**Extraction.** Covariate-related subnetworks are extracted by the greedy
algorithm conventional in dense subgraph discovery: the node of minimum degree
is removed at each iteration, and the node-induced subgraph maximizing the
objective over the resulting sequence is retained. Wu et al. (2022) define the
adaptive density function

```
f(S; lambda_W) = |W(S)| / |S| ^ lambda_W,     lambda_W in [1, 2]
```

for a node set `S` carrying total suprathreshold edge weight `|W(S)|`, which
interpolates between the two classical criteria: `lambda_W = 1` gives the
degree density `f1`, the objective of Charikar (2000), and `lambda_W = 2` gives
the area density `f2`. Equivalently, Chen et al. (2023) write the criterion as
an ℓ0 graph norm shrinkage penalty, `log||U||_1 - lambda_0 log||U||_0`, which
rewards edge weight within a subnetwork while penalizing its size.

`subnetr` parameterizes this family as

```
w / n ^ (2 * lambda)
```

for a set of `n` nodes carrying weight `w`. This is Chen et al.'s `lambda_0`
and half of Wu et al.'s `lambda_W`, so `lambda = 0.5` is degree density and
`lambda = 1` is area density. Larger values favour smaller, denser
subnetworks; `lambda = 0` places all nodes in a single subnetwork. Values
between 0.5 and 0.7 are the usual working range. Nodes removed in one round
are passed to the next, so subnetworks are extracted in decreasing order of
density, and nodes belonging to none of them are returned as background
singletons.

**Inference.** Testing several extracted subnetworks simultaneously requires
comparing subnetworks of different densities and sizes on a common scale.
Chen et al. (2023) address this with a concentration bound on the probability
of observing a subnetwork of size `v0` and density `γ` in a graph of overall
density `p`. `subnetr` instead uses the exact upper-tail binomial probability
of a subnetwork containing at least as many suprathreshold edges as observed,
given the overall suprathreshold edge probability, which serves the same
purpose without the bound's slack: subnetwork size enters as the binomial
sample size, so subnetworks of differing size are directly comparable.

The null distribution is obtained by permuting edge weights across the graph
and repeating the entire extraction, retaining the most extreme subnetwork
statistic from each permutation. This is a max-statistic (Westfall–Young)
procedure and controls the family-wise error rate over all extracted
subnetworks without further correction.

## Parameter tuning

Neither the screening threshold nor the size penalty should be fixed
arbitrarily. Wu et al. (2022) estimate `lambda` by maximizing a likelihood
under a stochastic block model, and treat the threshold `r` as a random
variable with prior `g(r)`, integrating the likelihood with respect to it
rather than selecting a single value. `subnetr` follows the same principle
over a grid, and the default is to do it for you.

### The default: supply nothing

`threshold` and `lambda` default to `NULL`, so

```r
fit <- subnet(W, n_perm = 999)
```

runs `tune_subnet()` over a grid of five `lambda` values by five thresholds,
fits at the selected pair, and — the part that matters — records that the
parameters were selected from the data, so the permutation null is built
accordingly.

Each candidate pair is scored by the log-likelihood ratio of a block model
against a homogeneous model, then standardized against that pair's own
permutation null. Standardization is what makes the grid comparable: a looser
threshold mechanically produces a denser graph, more blocks and a larger raw
likelihood, so the unstandardized statistic would simply select the loosest
setting on offer.

### Inspecting the choice

```{r tuning}
fit$tuning
```

Selection is on `z`, not `lr`. In the table above the second row has the higher
raw likelihood ratio but a lower `z`, precisely the mechanical effect that
standardization removes. The full grid is available as
`fit$tuning$grid`, a data frame with one row per candidate pair.

### Adjusting the grid

```r
fit <- subnet(W, tune = list(probs       = c(0.95, 0.98, 0.99),
                             lambda_grid = c(0.5, 0.6, 0.7),
                             n_perm      = 40))
```

`probs` are quantiles of the observed edge weights, defaulting to
`0.90, 0.95, 0.975, 0.99, 0.995`. The binding constraint is that screening must
retain at least as many edges as the target subnetwork spans: a 20-node
subnetwork in an 80-node connectome spans 190 of 3160 edges, about 6%, so a
99.5th-percentile threshold cannot recover it at any sample size. As a rule of
thumb the most stringent quantile in the grid should still retain a small
multiple of `c(c-1)/2` edges for the smallest subnetwork `c` of interest.

`lambda_grid` defaults to `seq(0.5, 0.9, 0.1)`, the range used in the source
papers. The `n_perm` here is the number of calibration permutations per grid
point and is unrelated to the `n_perm` of the test itself; 25 is enough to rank
candidates.

Setting `criterion = "likelihood"` skips the calibration permutations
altogether. It is faster, but biased toward the extremes of the threshold grid,
which is the reason calibration exists.

### Why tuning changes the null

Selecting the threshold from the same matrix that is subsequently tested is a
selection effect. Ignoring it approximately doubles the type-I error rate, so
that a nominal 5% test operates at about 10%. `subnet()` addresses this by
having each permutation repeat the same selection the observed data underwent:

| Null distribution | Type-I error at α = 0.05 |
|---|---|
| uncorrected | 0.10 |
| `subnetr` default (`"retune"`) | 0.04 |

This costs one extraction pass per grid point per permutation, which is
affordable because extraction is compiled.

### Reusing or fixing parameters

To inspect the tuning before fitting, or to reuse one search across several
analyses, pass the result object itself:

```r
tn  <- tune_subnet(W)
fit <- subnet(W, tuning = tn, n_perm = 999)
```

Pass the object, not its contents. The correction needs the grid that was
searched, and

```r
subnet(W, threshold = tn$threshold, lambda = tn$lambda)   # uncorrected
```

supplies the same two numbers with no record of where they came from, so the
null is built as though they had been chosen in advance. `subnet()` cannot
detect this, which is why `tuning =` exists.

When the parameters genuinely were fixed in advance — from prior work, say —
supplying them as numbers is correct, and the ordinary null is used
automatically:

```r
fit <- subnet(W, threshold = 2.5, lambda = 0.6, n_perm = 999)
```

## Performance and reproducibility

The screened graph is sparse by construction, and the compiled core is written
accordingly: peeling operates on a compressed sparse representation with a
lazily updated minimum-degree heap, and permutations place only the surviving
edges rather than reconstructing a dense matrix. Indicative single-core timings:

| Nodes | Edges | 200-permutation test |
|---|---|---|
| 100 | 4,950 | 0.01 s |
| 200 | 19,900 | 0.08 s |
| 400 | 79,800 | 0.41 s |

A 200-replicate power curve at 100 nodes completes in a few seconds.

Each permutation seeds its own random stream from its index, so results depend
only on `seed` and never on `n_cores` or on how the work is divided among
workers. Calling `set.seed()` before an unseeded call is sufficient for
reproducibility. Setting `n_cores > 1` uses OpenMP threads where the toolchain
provides them (see `has_openmp()`) and forked R workers otherwise.

## Documentation

```r
vignette("subnetr")                # the analysis pipeline in detail
vignette("power-analysis")         # study design
vignette("multiple-subnetworks")   # results with several subnetworks
```

## References

Chen S, Zhang Y, Wu Q, Bi C, Kochunov P, Hong LE (2023). Identifying
covariate-related subnetworks for whole-brain connectome analysis.
*Biostatistics*, kxad007.
[doi:10.1093/biostatistics/kxad007](https://doi.org/10.1093/biostatistics/kxad007)

Wu Q, Huang X, Culbreth AJ, Waltz JA, Hong LE, Chen S (2022). Extracting brain
disease-related connectome subgraphs by adaptive dense subgraph discovery.
*Biometrics* 78(4), 1566–1578.
[doi:10.1111/biom.13537](https://doi.org/10.1111/biom.13537)

## License

MIT
