---
title: "Multiple subnetworks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple subnetworks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9,
                      fig.height = 4.2, dpi = 120)
set.seed(1)
```

```{r}
library(subnetr)
```

## More than one subnetwork

Nothing in the method assumes a single subnetwork. A predictor may be
associated with several distinct systems at once, and `subnet()` returns each
of them separately, with its own p-value. This article covers what that looks
like, how multiplicity is handled across them, and where the approach runs out.

We plant two subnetworks of different sizes carrying different effects in a
90-node connectome. `simulate_fc()` recycles `f2` across `cluster_size`, so
each subnetwork gets its own effect size.

```{r simulate}
sim <- simulate_fc(n = 150, n_nodes = 90, cluster_size = c(20, 12),
                   f2 = c(0.08, 0.14), seed = 4)

lengths(sim$truth$nodes)
```

Node labels are shuffled, so neither block sits conveniently in the corner of
the matrix.

## Detection

```{r fit}
fit <- subnet(sim$W, n_perm = 999, seed = 2)
fit
```

Both planted subnetworks come back as separate blocks, and both are
significant. The remaining blocks are what greedy peeling always produces from
leftover noise — the densest block of a random graph is still a block — and
the test correctly declines to call them.

Checking against the truth:

```{r match}
for (k in seq_along(sim$truth$nodes)) {
  d <- vapply(fit$subnetworks, dice, numeric(1), b = sim$truth$nodes[[k]])
  best <- which.max(d)
  cat(sprintf("planted %d (size %2d): block #%d, Dice = %.2f, p = %.4f\n",
              k, length(sim$truth$nodes[[k]]), best, d[best],
              fit$p_value[best]))
}
```

The two panels below show why reordering matters. On the left the structure is
spread across arbitrary node indices and invisible; on the right both blocks
sit on the diagonal, outlined.

```{r plot}
plot(fit, what = "both")
```

## Blocks are disjoint by construction

Extraction partitions the node set. Each round pulls out the densest core and
hands the *peeled-off* nodes to the next round, so a node can only ever land in
one block.

```{r disjoint}
sn <- fit$subnetworks
sum(vapply(seq_along(sn), function(i) {
  if (i == length(sn)) return(0L)
  sum(vapply(sn[(i + 1):length(sn)], function(o) length(intersect(sn[[i]], o)),
             integer(1)))
}, integer(1)))

anyDuplicated(unlist(fit$partition$blocks))
```

Because every node has exactly one home, membership is a single label per node:

```{r membership}
table(membership(fit))
```

`0` is the background. `membership(fit, significant_only = FALSE)` labels every
extracted block instead of only the significant ones.

## Multiplicity is already handled

There is no extra correction to apply when several subnetworks are reported.
Each permutation reruns the **entire** extraction and records the single most
extreme block it produced anywhere in the graph; every observed block is then
compared against that distribution of maxima. This is the Westfall-Young
max-statistic construction, so the p-values control the family-wise error rate
jointly across all blocks returned — including the ones that fell short.

```{r fwer}
as.data.frame(fit)[, c("subnet", "size", "density", "p_value", "significant")]
```

Reporting two significant subnetworks out of seven extracted costs no
additional multiplicity budget. Do not pass these p-values through
`p.adjust()`.

## The null is conservative when signal is strong

Each block is tested against `pi0`, the supra-threshold edge rate across the
whole graph. That rate is computed from the observed data, so it *includes* the
signal edges, which makes it larger than the rate among genuinely null edges.

```{r conservative}
wv <- vech(fit$W)
K  <- sum(wv >= fit$threshold)   # supra-threshold edges, whole graph
M  <- length(wv)                 # all edges

sz <- fit$size[fit$significant]
inside_K <- sum(fit$n_edge[fit$significant])
inside_M <- sum(sz * (sz - 1) / 2)

c(pi0_used      = K / M,
  background    = (K - inside_K) / (M - inside_M),
  overstated_by = (K / M) / ((K - inside_K) / (M - inside_M)))
```

The test judges each block against a null roughly twice as dense as the real
background. That errs in the safe direction, but it does mean a weak second
subnetwork is harder to detect alongside a very strong first one than it would
be on its own. If you suspect a subtler subnetwork behind a dominant one,
rerunning on the remaining nodes gives it a cleaner null:

```{r residual, eval = FALSE}
rest <- setdiff(seq_len(nrow(sim$W)), fit$subnetworks[[1]])
subnet(sim$W[rest, rest], n_perm = 999)
```

Treat that as exploratory. The second pass conditions on the first result, so
its p-values are not jointly corrected with it.

## Power depends on subnetwork size

`power_curve()` accepts a vector `cluster_size` and reports each planted
subnetwork separately. Here both carry the *same* effect size, so any
difference in power is attributable to size alone.

```{r power}
pw <- power_curve(n = c(60, 100, 150), n_nodes = 90,
                  cluster_size = c(20, 12), f2 = c(0.03, 0.03),
                  n_sim = 50, n_perm = 199, seed = 5, progress = FALSE)

pw$by_cluster
```

The 20-node subnetwork is found long before the 12-node one. It spans 190
edges against 66, so it accumulates far more evidence at the same per-edge
effect size. Subnetwork size, not just effect size, drives what a study can
find.

The joint summary requires *every* planted subnetwork to be recovered, so it
tracks the harder of the two:

```{r joint}
pw$power[, c("n", "power", "power_recovery", "false_subnet_rate")]
```

`power` is the chance of declaring anything significant, `power_recovery` the
chance of recovering all planted subnetworks at `dice_cut` overlap, and
`false_subnet_rate` the share of significant blocks matching nothing planted.
When sizing a study aimed at several subnetworks, size it for the smallest one
you care about.

## Where this runs out

The partition is a genuine constraint, not a tuning choice: **overlapping
subnetworks cannot be represented.** If a hub region truly participates in two
distinct predictor-associated systems, peeling will assign it to whichever
block claims it first, and the other will be reported without it.

If overlap is central to your hypothesis, the block structure here is the wrong
model and you would be better served by a method that permits mixed membership.
If overlap is incidental — a node or two shared between otherwise distinct
systems — the cost is a slightly understated subnetwork rather than a wrong
answer.

## See also

`vignette("subnetr")` for the pipeline end to end, and
`vignette("power-analysis")` for study design with a single subnetwork.
