---
title: "Power analysis for subnetwork studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power analysis for subnetwork studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5,
                      fig.height = 4.5, dpi = 120)
set.seed(1)
```

```{r}
library(subnetr)
```

## Why simulate

There is no closed-form power calculation for this design. The test statistic
is the binomial tail probability of a block that a greedy search picked out by
maximizing density, compared against a permutation distribution of the same
quantity. Nothing about that is analytically tractable.

So we simulate the entire pipeline — generate a study, screen it, extract
subnetworks, run the permutation test — and count how often it works. That is
only practical because a full pipeline fit takes milliseconds; a curve over
five sample sizes with 120 replicates each is over 100,000 extractions and
runs in seconds.

## What you need to specify

A power analysis here is a statement about a scenario, and the scenario has
more moving parts than a two-sample t-test.

| argument | meaning | how to pick it |
|---|---|---|
| `n_nodes` | regions in the parcellation | you know this |
| `cluster_size` | nodes in the subnetwork you hope to find | the smallest scientifically interesting one |
| `f2` | Cohen's $f^2$ at an affected edge | from pilot data, or 0.02 / 0.15 / 0.35 for small / medium / large |
| `rho_in` | fraction of within-subnetwork edges actually affected | rarely 1; 0.8–0.9 is realistic |
| `rho_out` | fraction of outside edges affected | scattered associations; 0.01–0.05 |
| `n_cov` | nuisance covariates | age, sex, motion, site |

`f2` is the one people get wrong. At a single edge it is the effect size of the
predictor on that edge's connectivity, related to the partial correlation
$\rho$ by $f^2 = \rho^2 / (1 - \rho^2)$. A partial correlation of 0.14 — which
is a perfectly ordinary connectivity effect — is $f^2 = 0.02$.

```{r f2-table}
rho <- c(0.10, 0.14, 0.20, 0.30, 0.40)
data.frame(partial_r = rho, f2 = round(rho^2 / (1 - rho^2), 3))
```

## A first curve

```{r first-curve}
pw <- power_curve(n = c(60, 100, 140, 180, 220),
                  n_nodes = 100, cluster_size = 15, f2 = 0.02,
                  rho_in = 0.9, rho_out = 0.02,
                  n_sim = 120, n_perm = 199, seed = 7, progress = FALSE)
pw
```

```{r first-plot}
plot(pw, target = 0.8)
```

```{r required}
required_n(pw, target = 0.8)
```

### Two kinds of power

The two curves answer different questions and the gap between them is the
interesting part.

- **Any detection** — some subnetwork was declared significant. This is
  conventional power, and on its own it is a weak claim: it can be satisfied by
  a significant block that has nothing to do with the one you planted.
- **Recovery** — every planted subnetwork is matched by a significant one with
  Sørensen–Dice overlap of at least `dice_cut` (0.5 by default). This is the
  claim you actually want to make, because papers report *which regions* are
  involved.

A study in the gap between the two curves is large enough to detect that
something is there but too small to say where. If your conclusion names
regions, size the study on the recovery curve.

```{r gap}
with(pw$power, data.frame(n, any = power, recovery = power_recovery,
                          gap = round(power - power_recovery, 3)))
```

## The threshold dominates everything

Before trusting any number above, understand the setting that produces it.

Screening keeps the top few percent of edges. A subnetwork of `c` nodes spans
`c(c-1)/2` edges. If screening retains fewer edges *in the whole graph* than
the subnetwork contains, the subnetwork cannot survive as a block — no sample
size rescues it.

```{r threshold-sweep}
sweep <- lapply(c(0.99, 0.975, 0.95, 0.90), function(tp) {
  p <- power_curve(n = 120, n_nodes = 80, cluster_size = 20, f2 = 0.06,
                   n_sim = 60, n_perm = 199, seed = 7, progress = FALSE,
                   threshold_prob = tp)
  data.frame(threshold_prob = tp,
             edges_kept = round((1 - tp) * 80 * 79 / 2),
             recovery = p$power$power_recovery)
})
do.call(rbind, sweep)
```

The planted subnetwork here contains `20 * 19 / 2 = 190` edges. At the 99th
percentile only 32 edges survive screening in the entire graph, and recovery
collapses. Everything else about the study is unchanged.

`power_curve()` defaults `threshold_prob = NULL`, which scales the threshold
with the design so it retains roughly twice as many edges as the planted
subnetworks contain. That keeps the setting from silently dominating the
answer, but it does not make the dependence go away: **a power figure is a
statement about a particular screening choice and should be reported with it.**

## Being honest about tuning

The default fixes the threshold across replicates. If your real analysis will
instead tune it on the data — which is what `subnet()` does by default — the
simulation should pay the same selection cost, otherwise it reports power for a
procedure you are not going to run.

```{r tune-each}
honest <- power_curve(n = c(100, 180), n_nodes = 100, cluster_size = 15,
                      f2 = 0.02, n_sim = 40, n_perm = 199, seed = 7,
                      progress = FALSE, tune_each = TRUE,
                      tune = list(n_perm = 10,
                                  lambda_grid = c(0.5, 0.7, 0.9),
                                  probs = c(0.95, 0.975, 0.99)))
honest$power[, c("n", "power", "power_recovery")]
```

Shrinking the tuning grid via `tune` is the main lever for making this
affordable.

Note that tuning does not automatically come out *worse*. Two effects pull in
opposite directions: every permutation now has to search the same grid, which
costs power, but the threshold also adapts to each data set instead of being
fixed in advance, which gains power. In this scenario the second effect wins
and `tune_each = TRUE` reports slightly higher recovery than the fixed
default. Which way it goes depends on how well the fixed threshold happened to
suit the design.

The point of `tune_each` is not that it is conservative — it is that it
measures the procedure you will actually run. Quoting power from a fixed
threshold and then tuning at analysis time measures neither.

## Comparing designs

Because everything is fast, sweeping a design parameter is cheap. Does it pay
to increase the parcellation resolution?

```{r design-sweep}
grid <- expand.grid(n_nodes = c(60, 100, 200), n = c(100, 200))
res <- Map(function(nn, n) {
  p <- power_curve(n = n, n_nodes = nn, cluster_size = 15, f2 = 0.02,
                   n_sim = 60, n_perm = 199, seed = 11, progress = FALSE)
  data.frame(n_nodes = nn, n = n, recovery = p$power$power_recovery)
}, grid$n_nodes, grid$n)
do.call(rbind, res)
```

More nodes means more edges to search and a harsher effective multiplicity for
a subnetwork of fixed size, so at a fixed sample size a finer parcellation
costs power for a given absolute subnetwork size. Whether that is the right
trade depends on whether the subnetwork you care about is a fixed number of
regions or a fixed fraction of the brain.

## Cost and precision

Every reported proportion carries a Monte Carlo standard error, printed in
parentheses and available as `se_power` / `se_recovery`. It shrinks as
$1/\sqrt{n_{sim}}$, so halving it costs four times the compute.

```{r mc-error}
data.frame(n_sim = c(50, 100, 200, 500, 1000),
           se_at_50pct = round(sqrt(0.25 / c(50, 100, 200, 500, 1000)), 4))
```

Two hundred replicates give about ±0.035 at 50% power, which is enough to
choose between candidate sample sizes. Push to 1,000 only for a number going
into a grant.

Replicates are independent, so `n_cores` scales close to linearly, and results
do not depend on it:

```r
power_curve(n = c(60, 100, 140), n_nodes = 100, cluster_size = 15,
            f2 = 0.02, n_sim = 500, n_cores = 8, seed = 7)
```

## Reporting

A defensible power statement for this design names the scenario, not just the
number. For example:

> Assuming a 15-node subnetwork within a 100-region parcellation, an edge-level
> effect of $f^2 = 0.02$ (partial *r* ≈ 0.14) present in 90% of within-subnetwork
> edges, 2% of edges associated outside it, and screening at the 95th percentile
> of edge evidence, 200 subjects give 85% probability of recovering the
> subnetwork with Dice overlap ≥ 0.5 at a family-wise error rate of 0.05
> (200 simulation replicates, 199 permutations each; Monte Carlo SE 0.025).

## See also

`vignette("subnetr", package = "subnetr")` for the analysis pipeline itself.
