---
title: "Getting started with subnetr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with subnetr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 120)
set.seed(1)
```

```{r}
library(subnetr)
```

## The question

A connectome study measures functional connectivity between every pair of
regions and asks how it relates to something about the subject: a symptom
score, a drug dose, an age. With 100 regions there are 4,950 edges, and testing
each one separately means paying a multiplicity price that leaves almost
nothing standing.

`subnetr` asks a structured question instead. Rather than "which individual
edges are associated?", it asks "is there a **set of regions** whose mutual
connections are associated?" — a subnetwork. That is usually the scientific
question anyway, and it is far easier to answer, because a subnetwork
concentrates many weak edge-level signals into one strong block-level signal.

## The pipeline

Four steps, each a single function.

1. **`edge_stats()`** turns subject-level connectivity into one number per
   edge: the evidence that this edge relates to the predictor, on a
   $-\log_{10} p$ scale, adjusted for nuisance covariates.
2. **`tune_subnet()`** chooses how aggressively to screen those edges, and how
   the search should trade block size against block density.
3. **`subnet_extract()`** pulls dense blocks out of the screened graph by
   greedy peeling.
4. **`subnet()`** does 2–4 and attaches a permutation p-value to each block.

In practice you call `subnet()` and it runs the rest.

## Step 1: from data to a weighted graph

`edge_stats()` expects connectivity as an `n_subject` × `n_edge` matrix whose
rows are vectorized lower triangles, or as an `n_node` × `n_node` × `n_subject`
array.

```{r edge-stats}
set.seed(42)
n_subj  <- 120
n_nodes <- 60
n_edge  <- n_nodes * (n_nodes - 1) / 2

symptom <- rnorm(n_subj)
age     <- rnorm(n_subj, 40, 12)
motion  <- rexp(n_subj, 5)

fc <- matrix(rnorm(n_subj * n_edge), n_subj, n_edge)
fc <- fc + outer(motion, rnorm(n_edge, sd = 0.4))   # a nuisance effect

W <- edge_stats(fc, x = symptom, covariates = cbind(age, motion))
dim(W)
round(W[1:4, 1:4], 3)
```

Each edge is fitted as `y_e ~ symptom + age + motion`. Nothing here loops over
edges: the predictor and the connectivity matrix are residualized on the
covariates once, and the Frisch–Waugh–Lovell theorem then gives all 1,770
slopes, standard errors and t-statistics in a couple of matrix products.

P-values are computed on the log scale throughout, so evidence far past the
point where a p-value would underflow to zero is still represented exactly.

```{r attrs}
attr(W, "df")
range(attr(W, "t"))
```

## Step 2: simulating a study with known truth

To see the method work we need a data set where we know the answer.
`simulate_fc()` plants subnetworks with a specified effect size.

```{r simulate}
sim <- simulate_fc(n = 120, n_nodes = 60, cluster_size = c(15, 8),
                   f2 = c(0.15, 0.25), rho_in = 0.9, rho_out = 0.02,
                   seed = 2024)
sim
```

The arguments describe a scenario rather than a mechanism:

- `f2` is Cohen's $f^2$, the effect size at an affected edge.
- `rho_in = 0.9` means only 90% of the edges *inside* a planted subnetwork
  actually carry the effect. Real subnetworks are not uniformly affected, and a
  method that only works at `rho_in = 1` is not worth much.
- `rho_out = 0.02` scatters associated edges *outside* every subnetwork —
  the false positives any real screen has to cope with.
- Node labels are shuffled, so the planted structure is not conveniently
  sitting in the corner of the matrix.

`sim$truth` holds the answer for scoring:

```{r truth}
lengths(sim$truth$nodes)
head(sim$truth$nodes[[1]])
```

## Step 3: extraction

`subnet_extract()` screens the graph at a threshold and peels dense blocks out
of it.

```{r extract}
thr  <- quantile(vech(sim$W), 0.95)
part <- subnet_extract(sim$W, threshold = thr, lambda = 0.6)
part
```

The algorithm repeatedly deletes the lowest-degree node and scores whatever
remains after each deletion. The best-scoring set becomes a subnetwork; the
nodes deleted before it become the input to the next round. It stops when the
leftovers carry no supra-threshold weight.

### What `lambda` does

The score for a set of `n` nodes carrying total weight `w`, spanning
`m = n(n-1)/2` possible edges, is

$$(w/m)^\lambda \cdot w^{1-\lambda}.$$

At `lambda = 1` this is pure density, which favours small tight cliques. At
`lambda = 0` it is pure total weight, which favours large diffuse blocks.

```{r lambda-sweep}
sizes <- sapply(c(0.1, 0.3, 0.5, 0.7, 0.9), function(l) {
  subnet_extract(sim$W, threshold = thr, lambda = l)$sizes[1]
})
data.frame(lambda = c(0.1, 0.3, 0.5, 0.7, 0.9), top_block_size = sizes)
```

Two values of `lambda` are worth recognizing. At `lambda = 0.5` the score is
`w / n`, the average weighted degree, which is the classical densest-subgraph
criterion. As `lambda` approaches 1 the score approaches edge density, and the
extracted block shrinks toward the tightest small clique in the graph --
which is why `min_size` exists.

## Step 4: inference

Extraction on its own will always return *something* — the densest block of
pure noise is still a block. `subnet()` adds the test that decides whether it
means anything.

```{r subnet}
fit <- subnet(sim$W, n_perm = 999, seed = 1)
fit
```

Two ideas are doing the work.

**The block statistic puts different-sized blocks on one scale.** A block of
`n` nodes spans `m` possible edges, of which `k` cleared the threshold. If
edges were scattered at random at the graph's overall rate `pi0`, the chance of
seeing at least `k` is `pbinom(k - 1, m, pi0, lower.tail = FALSE)`. Because the
block's size enters as the binomial sample size, a small dense block and a
large moderately dense one are directly comparable.

**The null is generated by the same greedy search.** Each permutation
redistributes the edge weights at random, destroying subnetwork structure while
preserving the marginal distribution of weights, and then reruns the *whole*
extraction. Recording only the single most extreme block per permutation makes
this a Westfall–Young max-statistic procedure: the p-values already control
family-wise error across every block returned, and must not be adjusted again.

This also handles the selection problem that makes naive testing of a
discovered cluster meaningless. The blocks were chosen by maximizing density,
so of course they are dense; the null statistic is what that same maximization
produces when there is nothing to find.

```{r null-dist, fig.height = 3.6}
hist(fit$null_statistic / log(10), breaks = 40, col = "grey85", border = "white",
     xlab = expression(log[10]~"binomial tail probability"),
     main = "Permutation null (most extreme block per permutation)")
abline(v = fit$statistic[1] / log(10), col = "#d62728", lwd = 2)
```

The observed block sits far outside the null, which is why its p-value is at
the resolution floor of `1 / (n_perm + 1)`.

### Did it find the right thing?

```{r recovery}
sapply(sim$truth$nodes, function(tn) {
  max(sapply(fit$subnetworks[fit$significant], dice, b = tn))
})
```

`dice()` is the Sørensen–Dice overlap: 1 means the detected node set is exactly
the planted one.

```{r plot-fit, fig.height = 4.2, fig.width = 8}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fit, what = "observed")
plot(fit)
par(op)
```

The left panel is the matrix as measured; the right is the same matrix with
nodes reordered so the extracted blocks sit on the diagonal, with the
significant ones outlined. This plot is the fastest way to tell a genuine
compact subnetwork from a diffuse smear that happened to clear a threshold.

## Tuning, and why the null has to know about it

`subnet()` chose the threshold and `lambda` for you:

```{r tuning}
fit$tuning
```

Candidate settings are scored by the likelihood-ratio gain of a block model
over a homogeneous one. That statistic is not comparable across thresholds as
it stands, since changing the threshold changes the binary data being modelled.
`tune_subnet()` therefore standardizes each grid point against its own
permutation null, so grid points are compared on a common scale under common
random numbers.

Selecting parameters from the same matrix you then test creates a real problem.
Under the global null, testing at the selected values with a null built only at
those values roughly **doubles** the type-I error — a nominal 5% test runs at
about 11%. The fix is to let every permutation run the same search:

```{r null-modes}
fit$null
nrow(fit$null_grid)
```

Each permuted data set is scored at every grid point, picks its own
`(lambda, threshold)` by the same rule, and reports the maximum over the blocks
found there. Measured over 400 null replicates:

| null mode                   | nominal 0.05 | nominal 0.10 | nominal 0.20 |
|-----------------------------|--------------|--------------|--------------|
| `"selected"` (as published) | 0.100        | 0.177        | 0.315        |
| `"grid"` (conservative)     | 0.013        | 0.020        | 0.033        |
| `"retune"` (default)        | 0.035        | 0.087        | 0.207        |

If you fix `threshold` and `lambda` yourself, no selection happened and
`subnet()` uses the ordinary null automatically.

## Reproducibility

Every permutation seeds its own stream from its absolute index, so a result
depends on `seed` alone — not on `n_cores`, and not on how permutations were
split across workers.

```{r reproducible}
a <- subnet(sim$W, threshold = thr, lambda = 0.6, n_perm = 199, seed = 7)
b <- subnet(sim$W, threshold = thr, lambda = 0.6, n_perm = 199, seed = 7,
            n_cores = 2)
identical(a$p_value, b$p_value)
```

## Where to next

`vignette("power-analysis", package = "subnetr")` covers sizing a study:
turning an effect size and a target subnetwork into a sample size.
